---
title: "Descriptives and Graphs"
author: "Marilyn Piccirillo"
date: "9/10/2017"
output: pdf_document
---

```{r}
rm(list = ls())
library(foreign)
library(tidyverse)
library(plyr)
library(dplyr)
library(ggplot2)
library(lubridate)


#Pull in all three timepoints
SuicideT1 <- read.spss("StressStudy_T1short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT2 <- read.spss("StressStudy_T2short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT3 <- read.spss("StressStudy_T3short.sav", use.value.labels = FALSE, to.data.frame = TRUE)

#Merge all three timepoints
SuicideMerge <- left_join(SuicideT1, SuicideT2, by = "ID")
SuicideMerge <- left_join(SuicideMerge, SuicideT3, by= "ID")
#Change order of columns
SuicideMerge <- SuicideMerge[, c(1:2, 22, 28, 16:21, 23:27, 29:33, 3:15)]
```

Move your data into a long format and a wide format. Did you have any specific challenges that you encountered? If so, discuss them.
```{r}
#Data is originally in wide mode. Convert to long format:
SuicideLong <- gather(SuicideMerge, key = "Time", value = "BSS", select(-ID, T1_BSSTot, T2_BSSTot, T3_BSSTot)) %>%
arrange(SuicideLong$ID)

SuicideLong

#Convert back to wide mode:
long_to_wide <- wide_to_long %>% 
  spread(Weeks, Value)
long_to_wide
```
I was inititally very confused by the purpose of the "Value" column - it took me a while to realize that this column wolud contain all the data for each variable in the long format. Upon converting back to a wide format, I realized that it arranges the variables in alphabetical order...not sure if that is ideal. But it is easy enough to re-arrange variables if necessary.

Create a wave variable and date variable (if applicable).
```{r}
#Break apart date variables
SuicideMerge$T1_Date <- as.Date(SuicideMerge$T1_DateStarted)
SuicideMerge$T2_Date <- as.Date(SuicideMerge$T2_DateStarted)
SuicideMerge$T3_Date <- as.Date(SuicideMerge$T3_DateStarted)
#Change order of columns
#SuicideMerge <- SuicideMerge[, c(1, 34:36, 5:33, 2:4)]
```

What is your sample size for each wave of assessment?
```{r}
length(which(!is.na(SuicideMerge$T1_Date))) #553 in the first wave
length(which(!is.na(SuicideMerge$T2_Date))) #278 in the second wave
length(which(!is.na(SuicideMerge$T3_Date))) #138 in the third wave
```

Take the date variable and convert it to a different date format such as time in study or age (if appropriate). What scale is most suitable for your analyses? (weeks/months/years?)
```{r}
#Create a difference variable for time (Days)
SuicideMerge$Days <- difftime(SuicideMerge$T3_Date , SuicideMerge$T1_Date , units = c("days"))
#Change order of columns
#SuicideMerge <- SuicideMerge[, c(1, 37, 2:36)]

#Create a difference variable for time (Weeks)
SuicideMerge$Weeks <- difftime(SuicideMerge$T3_Date , SuicideMerge$T1_Date , units = c("weeks"))
SuicideMerge$Weeks <- round(SuicideMerge$Weeks, digits = 2)
#SuicideMerge <- SuicideMerge[, c(1, 38, 2:37)]
```
Weeks is probably the most suitable for my analyses.

Graph your data using the different time metrics, fitting individual curves for each person.
```{r}
SuicideLong <- SuicideMerge %>% 
  gather(2, 8, 14, 19, key = "Weeks", value = "BSSTot") %>% 
  arrange(ID)

Graph1 <- ggplot(SuicideMerge,
   aes(x = "Week", y = "T1_BSSTot", group = ID)) + geom_point()  
print(Graph1)
```


Create an overall average trend of your data (split up into groups if appropriate). Attempt to color your individual data points and/or shade different lines (highlight some particiapnts, highlight the average trend line but not the individual level lines)
```{r}
AvgTrend <-  gg1 <- ggplot(SuicideMerge, aes(x = SuicideMerge$Week, y = SMN7)) + geom_point() + stat_smooth() 
gg7
```


Look at the correlations of your DV across time
```{r}
cor(SuicideMerge$T1_BSSTot, SuicideMerge$T2_BSSTot, use = "pairwise.complete.obs")
cor(SuicideMerge$T2_BSSTot, SuicideMerge$T3_BSSTot, use = "pairwise.complete.obs")
```
Correlation between suicide score at T1 and T2 = .65; at T2 and T3 = .46
