---
title: "Descriptives and Graphs"
author: "Marilyn Piccirillo"
date: "9/10/2017"
output: pdf_document
---

```{r}
rm(list = ls())
library(foreign)
library(tidyverse)
library(plyr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(lubridate)


#Pull in all three timepoints
SuicideT1 <- read.spss("StressStudy_T1short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT2 <- read.spss("StressStudy_T2short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT3 <- read.spss("StressStudy_T3short.sav", use.value.labels = FALSE, to.data.frame = TRUE)

#Merge all three timepoints
SuicideMerge <- left_join(SuicideT1, SuicideT2, by = "ID")
SuicideMerge <- left_join(SuicideMerge, SuicideT3, by= "ID")
#Change order of columns
SuicideMerge <- SuicideMerge[, c(1:2, 22, 28, 16:21, 23:27, 29:33, 3:15)]

SuicideMerge$T1_Date <- as.Date(SuicideMerge$T1_DateStarted)
SuicideMerge$T2_Date <- as.Date(SuicideMerge$T2_DateStarted)
SuicideMerge$T3_Date <- as.Date(SuicideMerge$T3_DateStarted)

SuicideShort <- select(SuicideMerge, ID, T1_Date, T2_Date, T3_Date, T1_BSSTot, T2_BSSTot, T3_BSSTot)
SuicideShort <- as.tibble(SuicideShort)
```

Move your data into a long format and a wide format. Did you have any specific challenges that you encountered? If so, discuss them.
```{r}
#Data is originally in wide mode. Convert to long format:
Suicidelong <- gather (SuicideShort, Timepoint, BSSscore, T1_BSSTot:T3_BSSTot)
Suicidelong <- as.tibble(Suicidelong) 
Suicidelong <- arrange(Suicidelong, .by_group = ID)

#Convert back to wide mode:
Suicidewide <- spread(Suicidelong, Timepoint, BSSscore)
Suicidewide <- as.tibble(Suicidewide)
```
Identifying the key variable was pretty confusing! It was easier when I reduced the dataset down to 4 variables...

Create a wave variable and date variable (if applicable).
```{r}
#Break apart date variables
#Code copied from above
SuicideMerge$T1_Date <- as.Date(SuicideMerge$T1_DateStarted)
SuicideMerge$T2_Date <- as.Date(SuicideMerge$T2_DateStarted)
SuicideMerge$T3_Date <- as.Date(SuicideMerge$T3_DateStarted)
#Change order of columns
#SuicideMerge <- SuicideMerge[, c(1, 34:36, 5:33, 2:4)]
```

What is your sample size for each wave of assessment?
```{r}
length(which(!is.na(SuicideMerge$T1_Date))) #553 in the first wave
length(which(!is.na(SuicideMerge$T2_Date))) #278 in the second wave
length(which(!is.na(SuicideMerge$T3_Date))) #138 in the third wave
```

Take the date variable and convert it to a different date format such as time in study or age (if appropriate). What scale is most suitable for your analyses? (weeks/months/years?)
```{r}
#Create a difference variable for time (Days)
SuicideShort$Days <- difftime(SuicideShort$T3_Date , SuicideShort$T1_Date , units = c("days"))

#Create a difference variable for time (Weeks)
SuicideShort$Weeks <- difftime(SuicideShort$T3_Date , SuicideShort$T1_Date , units = c("weeks"))
SuicideShort$Weeks <- round(SuicideShort$Weeks, digits = 2)
```
Weeks is probably the most suitable for my analyses.

Graph your data using the different time metrics, fitting individual curves for each person.
```{r}
Suicidelong <- gather (SuicideShort, Timepoint, BSSscore, T1_BSSTot:T3_BSSTot)
Suicidelong <- as.tibble(Suicidelong) 
Suicidelong <- arrange(Suicidelong, .by_group = ID)
Suicidelong$Timepoint <- as.factor(Suicidelong$Timepoint)

gg1 <- ggplot(Suicidelong,
   aes(x = Timepoint, y = BSSscore, group = ID)) + geom_line() +  geom_point()
gg1
```


Create an overall average trend of your data (split up into groups if appropriate). Attempt to color your individual data points and/or shade different lines (highlight some particiapnts, highlight the average trend line but not the individual level lines)
```{r}
Avgtrendgg <-  ggplot(Suicidelong,
                      aes(x = Timepoint, y = BSSscore)) + geom_point() + stat_smooth() 
Avgtrendgg
```


Look at the correlations of your DV across time
```{r}
cor(SuicideMerge$T1_BSSTot, SuicideMerge$T2_BSSTot, use = "pairwise.complete.obs")
cor(SuicideMerge$T2_BSSTot, SuicideMerge$T3_BSSTot, use = "pairwise.complete.obs")
```
Correlation between suicide score at T1 and T2 = .65; at T2 and T3 = .46
