---
title: "Final Paper Analyses"
author: "Marilyn Piccirillo"
date: "12/12/2017"
output: pdf_document
---
```{r, echo = FALSE}
rm(list = ls())
library(foreign)
library(psych)
library(tidyverse)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(lubridate)
library(lme4)
library(lavaan)
library(sjstats)
library(sjPlot)
library(semPlot)

#Pull in all three timepoints
DepressionT1 <- read.spss("StressStudy_T1short_rename.sav", use.value.labels = FALSE, to.data.frame = TRUE)
DepressionT2 <- read.spss("StressStudy_T2short_rename.sav", use.value.labels = FALSE, to.data.frame = TRUE)
DepressionT3 <- read.spss("StressStudy_T3short_rename.sav", use.value.labels = FALSE, to.data.frame = TRUE)

#Merge all three timepoints
DepressionMerge <- left_join(DepressionT1, DepressionT2, by = "ID")
DepressionMerge <- left_join(DepressionMerge, DepressionT3, by= "ID")

#Make a shorter dataset
#Depressionshort <- dplyr::select(DepressionMerge, ID, X._created_at.x, X._created_at.y, X._created_at, T1_BDI, T2_BDI, T3_BDI, T1_SIAS, T2_SIAS, T3_SIAS, T1_ACSS, T2_ACSS, T3_ACSS, T1_Burden, T1_Belong, T1_INQ, T2_Burden, T2_Belong, T2_INQ, T3_Burden, T3_Belong, T3_INQ, q_gender_0_score, q_age_0, q_ethnicity_score, q_hispanic_score, q_sexuality_score)


#Converting into proper format
 DepressionMerge$Gender <- as.factor(DepressionMerge$Gender)
 DepressionMerge$Ethnicity <- as.factor(DepressionMerge$Ethnicity)
 DepressionMerge$Hispanic <- as.factor(DepressionMerge$Hispanic)
 DepressionMerge$Sexuality <- as.factor(DepressionMerge$Sexuality)

#Separatedates
DepressionMerge$T1_Date <- as.Date(DepressionMerge$T1_Date)
DepressionMerge$T2_Date <- as.Date(DepressionMerge$T2_Date)
DepressionMerge$T3_Date <- as.Date(DepressionMerge$T3_Date)

#Make date variables
DepressionMerge$T1_Day <- difftime(DepressionMerge$T1_Date, DepressionMerge$T3_Date, units = c("days"))
DepressionMerge$T2_Day <- difftime(DepressionMerge$T2_Date, DepressionMerge$T3_Date, units = c("days"))
DepressionMerge$T3_Day <- difftime(DepressionMerge$T3_Date, DepressionMerge$T3_Date, units = c("days"))

#Make Wave variables
DepressionMerge$Wave1 <- rep(1, 553)
DepressionMerge$Wave2 <- rep(2, 553)
DepressionMerge$Wave3 <- rep(3, 553)

#Shorten and rearrange Depressionshort
Depressionshort <- dplyr::select(DepressionMerge, ID, Wave1, Wave2, Wave3, T1_BDI, T2_BDI, T3_BDI, T1_SIAS, T2_SIAS, T3_SIAS, T1_Belong, T1_Burden, T2_Belong, T2_Burden, T3_Belong, T3_Burden, T1_ACSS, T2_ACSS, T3_ACSS)

#Selecting only people who completed at least two time points 
#DepressionMerge <- subset(DepressionMerge, T1_ACSS >= "0" & T2_ACSS >= "0" | 
#T3_ACSS >= "0")

rm(DepressionT1, DepressionT2, DepressionT3)
```

Testing a longitudinal path model to replicate the model proposed by Davidson et al. (2011). They examined how interpersonal needs (Percieved burdensomness and thwarted belongingness) mediated the relationship of depression and social anxiety predicting suicidality. The issue with their model is that they used cross-sectional data. I'd like to replicate their model using longitundial data. There weren't enough people who indicated symptoms of suicidality, so I will use the Acquired Capability for Suicide Scale, which assesses constructs such as fearlessness towards death. ACSS scores are highly correlated with suicidality.

I will test the following mediation relationships:

Depression --> Percieved Burdensomness, Thwarted Belongingness --> Acquired Capability for Suicide
Social Anxiety --> Percieved Burdensomness, Thwarted Belongingness --> Acquired Capability for Suicide

I expect that the mediation effect will be stronger for Depression --> Percieved burdensomness --> ACSS and that the mediation effect will be stronger for Social Anxiety --> Thwarted Belongingness --> ACSS

```{r}

medDep <- '

#Correlate residuals

T1_BDI ~~ T2_BDI
T2_BDI ~~ T3_BDI

T1_Belong ~~ T2_Belong
T2_Belong ~~ T3_Belong

T1_Burden ~~ T2_Burden
T2_Burden ~~ T3_Burden

T1_ACSS ~~ T2_ACSS
T2_ACSS ~~ T3_ACSS

#Regressions to calculate indirect effects

T1_Burden ~ a1*T1_BDI + c1*T1_ACSS
T1_Belong ~ b1*T1_BDI + d1*T1_ACSS

T2_Belong ~ a1*T1_BDI + c2*T2_ACSS + g1*T1_Belong
T2_Burden ~ b1*T1_BDI + d2*T2_ACSS + h1*T2_Burden

T3_Belong ~ a2*T2_BDI + c3*T3_ACSS + g2*T2_Belong
T3_Burden ~ b2*T2_BDI + d3*T3_ACSS + h2*T2_Burden

T2_ACSS ~ i1*T1_ACSS + e1*T1_BDI + c1*T1_Belong + d1*T1_Burden
T3_ACSS ~ i2*T2_ACSS + e1*T1_BDI + c2*T2_Belong + d2*T2_Belong

#BDI --> Belong (a)
#BDI --> Burden (b)
#Belong --> ACSS (c)
#Burden --> ACSS (d)
#BDI --> ACSS (e)
#AR: BDI--> BDI (f)
#AR: Belong --> Belong (g)
#AR: Burden --> Burden (h)
#AR: ACSS --> ACSS (i)

ind:= a1*i2*c2*g1*g2 + b1*i2*d2*h1*h2 
total:= ind + e1 + i2 '

fit.medDep <- sem(medDep, data = Depressionshort, missing = "ml")
summary(fit.medDep, standardized = TRUE, fit.measures = TRUE) 
```
```{r}
medSIAS <- '


#Correlate residuals

T1_SIAS ~~ T2_SIAS
T2_SIAS ~~ T3_SIAS

T1_Belong ~~ T2_Belong
T2_Belong ~~ T3_Belong

T1_Burden ~~ T2_Burden
T2_Burden ~~ T3_Burden

T1_ACSS ~~ T2_ACSS
T2_ACSS ~~ T3_ACSS

#Regressions to calculate indirect effects

T1_Burden ~ a1*T1_SIAS + c1*T1_ACSS
T1_Belong ~ b1*T1_SIAS + d1*T1_ACSS

T2_Belong ~ a1*T1_SIAS + c2*T2_ACSS + g1*T1_Belong
T2_Burden ~ b1*T1_SIAS + d2*T2_ACSS + h1*T2_Burden

T3_Belong ~ a2*T2_SIAS + c3*T3_ACSS + g2*T2_Belong
T3_Burden ~ b2*T2_SIAS + d3*T3_ACSS + h2*T2_Burden

T2_ACSS ~ i1*T1_ACSS + e1*T1_SIAS + c1*T1_Belong + d1*T1_Burden
T3_ACSS ~ i2*T2_ACSS + e1*T1_SIAS + c2*T2_Belong + d2*T2_Belong

#SIAS --> Belong (a)
#SIAS --> Burden (b)
#Belong --> ACSS (c)
#Burden --> ACSS (d)
#SIAS --> ACSS (e)
#AR: SIAS--> SIAS (f)
#AR: Belong --> Belong (g)
#AR: Burden --> Burden (h)
#AR: ACSS --> ACSS (i)

ind:= a1*i2*c2*g1*g2 + b1*i2*d2*h1*h2 
total:= ind + e1 + i2 '

fit.medSIAS <- sem(medSIAS, data = Depressionshort, missing = "ml")
summary(fit.medSIAS, standardized = TRUE, fit.measures = TRUE) 
```

```{r}
#Descriptive stats
table(DepressionMerge$Gender)
table(DepressionMerge$Ethnicity)
table(DepressionMerge$Hispanic)
describe(DepressionMerge$Age)
```

```{r, eval = FALSE}
#Subset data to calculate alphas
T1BDI <- dplyr::select(DepressionMerge, T1_BDI1:T1_BDI21)
T2BDI <- dplyr::select(DepressionMerge, T2_BDI1:T2_BDI21)
T3BDI <- dplyr::select(DepressionMerge, T3_BDI1:T3_BDI21)
T1SIAS <- dplyr::select(DepressionMerge, T1_SIAS1:T1_SIAS4, T1_SIAS6:T1_SIAS8, T1_SIAS10, T1_SIAS12:T1_SIAS20)
T2SIAS <- dplyr::select(DepressionMerge, T2_SIAS1:T2_SIAS4, T2_SIAS6:T2_SIAS8, T2_SIAS10, T2_SIAS12:T2_SIAS20)
T3SIAS <- dplyr::select(DepressionMerge, T3_SIAS1:T3_SIAS4, T3_SIAS6:T3_SIAS8, T3_SIAS10, T3_SIAS12:T3_SIAS20)
T1ACSS <- dplyr::select(DepressionMerge, T1_ACSS1:T1_ACSS5)
T2ACSS <- dplyr::select(DepressionMerge, T2_ACSS1:T2_ACSS5)
T3ACSS <- dplyr::select(DepressionMerge, T3_ACSS1:T3_ACSS5)
T1Belong <- dplyr::select(DepressionMerge, T1_INQ7:T1_INQ15)
T2Belong <- dplyr::select(DepressionMerge, T2_INQ7:T2_INQ16)
T3Belong <- dplyr::select(DepressionMerge, T3_INQ7:T3_INQ16)
T1Burden <- dplyr::select(DepressionMerge, T1_INQ1:T1_INQ6)
T2Burden <- dplyr::select(DepressionMerge, T2_INQ1:T2_INQ6)
T3Burden <- dplyr::select(DepressionMerge, T3_INQ1:T3_INQ6)

alpha(T1BDI)
alpha(T2BDI)
alpha(T3BDI)
alpha(T1SIAS)
alpha(T2SIAS)
alpha(T3SIAS)
alpha(T1ACSS, check.keys = TRUE)
alpha(T2ACSS, check.keys = TRUE)
alpha(T3ACSS, check.keys = TRUE)
alpha(T1Burden)
alpha(T2Burden)
alpha(T3Burden)
alpha(T1Belong, check.keys = TRUE)
alpha(T2Belong, check.keys = TRUE)
alpha(T3Belong, check.keys = TRUE)
```
