---
title: "Growth curves 11.30.17"
author: "Marilyn Piccirillo"
date: "11/30/2017"
output: pdf_document
---

```{r}
rm(list = ls())
library(foreign)
library(tidyverse)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(lubridate)
library(lme4)
library(lavaan)
library(sjstats)
library(sjPlot)
library(semPlot)

#Pull in all three timepoints
DepressionT1 <- read.spss("StressStudy_T1short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
DepressionT2 <- read.spss("StressStudy_T2short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
DepressionT3 <- read.spss("StressStudy_T3short.sav", use.value.labels = FALSE, to.data.frame = TRUE)

#Pull in all three timepoints for SEM measurement models.
DepressionT1SEM <- read.spss("Stress Study Database_T1_useforSEM.sav", use.value.labels = FALSE, to.data.frame = TRUE)
DepressionT1SEM <- plyr::rename(DepressionT1SEM, c("q_bdi_1_score" = "T1BDI1", "q_bdi_2_score" = "T1BDI2", "q_bdi_3_score" = "T1BDI3", "q_bdi_4_score" = "T1BDI4", "q_bdi_5_score" = "T1BDI5", "q_bdi_6_score" = "T1BDI6", "q_bdi_7_score" = "T1BDI7", "q_bdi_8_score" = "T1BDI8", "q_bdi_9_score" = "T1BDI9", "q_bdi_10_score" = "T1BDI10", "q_bdi_11_score" = "T1BDI11", "q_bdi_12_score" = "T1BDI12", "q_bdi_13_score" = "T1BDI13", "q_bdi_14_score" = "T1BDI14", "q_bdi_15_score" = "T1BDI15", "q_bdi_16_score" = "T1BDI16", "q_bdi_17_score" = "T1BDI17", "q_bdi_18_score" = "T1BDI18", "q_bdi_19_score" = "T1BDI19", "q_bdi_20_score" = "T1BDI20", "q_bdi_21_score" = "T1BDI21"))
DepressionT2SEM <- read.spss("Stress Study Database_T2_useforSEM.sav", use.value.labels = FALSE, to.data.frame = TRUE)
DepressionT2SEM <- plyr::rename(DepressionT2SEM, c("q_bdi_1_score" = "T2BDI1", "q_bdi_2_score" = "T2BDI2", "q_bdi_3_score" = "T2BDI3", "q_bdi_4_score" = "T2BDI4", "q_bdi_5_score" = "T2BDI5", "q_bdi_6_score" = "T2BDI6", "q_bdi_7_score" = "T2BDI7", "q_bdi_8_score" = "T2BDI8", "q_bdi_9_score" = "T2BDI9", "q_bdi_10_score" = "T2BDI10", "q_bdi_11_score" = "T2BDI11", "q_bdi_12_score" = "T2BDI12", "q_bdi_13_score" = "T2BDI13", "q_bdi_14_score" = "T2BDI14", "q_bdi_15_score" = "T2BDI15", "q_bdi_16_score" = "T2BDI16", "q_bdi_17_score" = "T2BDI17", "q_bdi_18_score" = "T2BDI18", "q_bdi_19_score" = "T2BDI19", "q_bdi_20_score" = "T2BDI20", "q_bdi_21_score" = "T2BDI21"))
DepressionT3SEM <- read.spss("Stress Study Database_T3_useforSEM.sav", use.value.labels = FALSE, to.data.frame = TRUE)
DepressionT3SEM <- plyr::rename(DepressionT3SEM, c("q_bdi_1_score" = "T3BDI1", "q_bdi_2_score" = "T3BDI2", "q_bdi_3_score" = "T3BDI3", "q_bdi_4_score" = "T3BDI4", "q_bdi_5_score" = "T3BDI5", "q_bdi_6_score" = "T3BDI6", "q_bdi_7_score" = "T3BDI7", "q_bdi_8_score" = "T3BDI8", "q_bdi_9_score" = "T3BDI9", "q_bdi_10_score" = "T3BDI10", "q_bdi_11_score" = "T3BDI11", "q_bdi_12_score" = "T3BDI12", "q_bdi_13_score" = "T3BDI13", "q_bdi_14_score" = "T3BDI14", "q_bdi_15_score" = "T3BDI15", "q_bdi_16_score" = "T3BDI16", "q_bdi_17_score" = "T3BDI17", "q_bdi_18_score" = "T3BDI18", "q_bdi_19_score" = "T3BDI19", "q_bdi_20_score" = "T3BDI20", "q_bdi_21_score" = "T3BDI21"))

DepT1 <- left_join(DepressionT1, DepressionT1SEM, by = "ID")
DepT2 <- left_join(DepressionT2, DepressionT2SEM, by = "ID")
DepT3 <- left_join(DepressionT3, DepressionT3SEM, by = "ID")
DepressionSEM <- left_join(DepT1, DepT2, by = "ID")
DepressionSEM <- left_join(DepressionSEM, DepT3, by = "ID")

#Separatedates
DepressionSEM$T1_Date <- as.Date(DepressionSEM$T1_DateStarted)
DepressionSEM$T2_Date <- as.Date(DepressionSEM$T2_DateStarted)
DepressionSEM$T3_Date <- as.Date(DepressionSEM$T3_DateStarted)

#Make a shorter dataset
DepressionSEM <- dplyr::select(DepressionSEM, ID, T1_Date, T2_Date, T3_Date, Gender, Age, Ethn, SexOrien, T1_BSSTot, T2_BSSTot, T3_BSSTot, T1_BDITot, T2_BDITot, T3_BDITot, T1_SIASTot, T2_SIASTot, T3_SIASTot, T1_ACSSTot, T2_ACSSTot, T3_ACSSTot, T1BDI1:T1BDI21, T2BDI1:T2BDI21, T3BDI1:T3BDI21)
#Converting into proper format
DepressionSEM$Gender <- as.factor(DepressionSEM$Gender)
DepressionSEM$Ethn <- as.factor(DepressionSEM$Ethn)
DepressionSEM$SexOrien <- as.factor(DepressionSEM$SexOrien)
DepressionSEM$Age <- as.character(DepressionSEM$Age)
DepressionSEM$Age <- as.numeric(DepressionSEM$Age)
DepressionSEM <- as.tibble(DepressionSEM)

#Make date variables
DepressionSEM$T1_Day <- difftime(DepressionSEM$T1_Date, DepressionSEM$T1_Date, units = c("days"))
DepressionSEM$T2_Day <- difftime(DepressionSEM$T2_Date, DepressionSEM$T1_Date, units = c("days"))
DepressionSEM$T3_Day <- difftime(DepressionSEM$T3_Date, DepressionSEM$T1_Date, units = c("days"))

#Depressionwide <- select(Depressionshort, everything())
#Selecting only people who completed all three time points
DepressionSEMShort <- subset(DepressionSEM, T1_BDITot >= "0" & (T2_BDITot >= "0" | T3_BDITot >= "0"))

#Convert DepressionShort to longform
#Depressionlong <- gather(Depressionshort, Var, Val, select = c("T1_ACSSTot", "T1_BDITot", "T1_BSSTot", "T1_SIASTot", "T2_ACSSTot", "T2_BDITot", "T2_BSSTot", "T2_SIASTot", "T3_ACSSTot", "T3_BDITot", "T3_BSSTot", "T3_SIASTot", "T1_Day", "T2_Day", "T3_Day"))
#Depressionlong <- arrange(Depressionlong, .by_group = ID)
#Depressionlong <- separate (Depressionlong, Var, c("Timepoint", "Variable"), sep = "_")
#Depressionlong <- spread(Depressionlong, Variable, Val)
#Depressionlong <- as.tibble(Depressionlong) 
#Removing someone with bad T2 data
#Depressionlong <- Depressionlong[-c(784:786), ]

#Depressionlong <- dplyr::select(Depressionlong, -Timepoint)

#Create a wave variable...repeat 1,2,3 for each person
#n<- c(1, 2, 3)
#DepressionSEMShort$Wave <- rep(n, 273)
#Make a shorter dataset to work with.
#Depressionlong <- dplyr::select(Depressionlong, ID, Wave, Day, Gender, Age, Ethn, SexOrien, BDITot, BSSTot, SIASTot, ACSSTot)

#Make Wave variables
DepressionSEMShort$Wave1 <- rep(1, 273)
DepressionSEMShort$Wave2 <- rep(2, 273)
DepressionSEMShort$Wave3 <- rep(3, 273)

#Shorten and rearrange Depressionshort
#Depressionshort <- dplyr::select(Depressionshort, ID, Wave1, Wave2, Wave3, T1_Day, T2_Day, T3_Day, Gender, Age, Ethn, SexOrien, T1_BDITot, T2_BDITot, T3_BDITot, T1_BSSTot, T2_BSSTot, T3_BSSTot, T1_SIASTot, T2_SIASTot, T3_SIASTot, T1_ACSSTot, T2_ACSSTot, T3_ACSSTot)

#Clean up
rm(DepressionT1, DepressionT2, DepressionT3, DepressionT1SEM, DepressionT2SEM, DepressionT3SEM, DepT1, DepT2, DepT3)
```

Test measurement invariance across time for your construct. Can you run growth models? If there is evidence of non-invariance, what seems to be the problem?

Configural
```{r}
#For latent variables, I will use cardinal symptoms of depression: low mood, loss of pleasure (BDI1, BDI2, BDI4)

config <- '
Dep1 =~ T1BDI1 + T1BDI2 + T1BDI4
Dep2 =~ T2BDI1 + T2BDI2 + T2BDI4
Dep3 =~ T3BDI1 + T3BDI2 + T3BDI4

T1BDI1 ~~ T1BDI2 + T1BDI4
T1BDI2 ~~ T1BDI4
T2BDI1 ~~ T2BDI2 + T2BDI4
T2BDI2 ~~ T2BDI4
T3BDI1 ~~ T3BDI2 + T3BDI4
T3BDI2 ~~ T3BDI4
'

config <- cfa(config, data = DepressionSEMShort, meanstructure = TRUE, std.lv = TRUE)
summary(config, standardized = TRUE, fit.measures = TRUE)
#Recieved warnings about standard errors, model may not be identified or variances are negative. "cov.lv" doesn't reveal any negative variances
```
Constrain - Weak
```{r}
weak <- '
Dep1 =~ L1*T1BDI1 + L2*T1BDI2 + L3*T1BDI4
Dep2 =~ L1*T2BDI1 + L2*T2BDI2 + L3*T2BDI4
Dep3 =~ L1*T3BDI1 + L2*T3BDI2 + L3*T3BDI4

Dep2 ~~ NA*Dep2
Dep3 ~~ NA*Dep3

T1BDI1 ~~ T1BDI2 + T1BDI4
T1BDI2 ~~ T1BDI4
T2BDI1 ~~ T2BDI2 + T2BDI4
T2BDI2 ~~ T2BDI4
T3BDI1 ~~ T3BDI2 + T3BDI4
T3BDI2 ~~ T3BDI4
'

weak <- cfa(weak, data = DepressionSEMShort, meanstructure = TRUE, std.lv = TRUE)
summary(weak, standardized = TRUE, fit.measures = TRUE)
#Reciving warnings about matrix not being positive definite. "cov.lv" didn't show any negative variances; "theta" showed a few negative theta values....

anova(config, weak)
#No difference between the configural model and the weak - constrained model.
```


Constrain - Strong
```{r}
strong <- '
Dep1 =~ L1*T1BDI1 + L2*T1BDI2 + L3*T1BDI4
Dep2 =~ L1*T2BDI1 + L2*T2BDI2 + L3*T2BDI4
Dep3 =~ L1*T3BDI1 + L2*T3BDI2 + L3*T3BDI4

Dep2 ~~ NA*Dep2
Dep3 ~~ NA*Dep3

T1BDI1 ~~ T1BDI2 + T1BDI4
T1BDI2 ~~ T1BDI4
T2BDI1 ~~ T2BDI2 + T2BDI4
T2BDI2 ~~ T2BDI4
T3BDI1 ~~ T3BDI2 + T3BDI4
T3BDI2 ~~ T3BDI4

T1BDI1 ~ t1*1
T1BDI2 ~ t2*1
T1BDI4 ~ t3*1

T2BDI1 ~ t1*1
T2BDI2 ~ t2*1
T2BDI4 ~ t3*1

T3BDI1 ~ t1*1
T3BDI2 ~ t2*1
T3BDI4 ~ t3*1

Dep2 ~ NA*1
Dep3 ~ NA*1'

strong <- cfa(strong, data = DepressionSEMShort, meanstructure = TRUE, std.lv = TRUE)
summary(strong, standardized = TRUE, fit.measures = TRUE)
#Again, warnings - "cov.lv" looks fine, "theta" shows some negative values
anova(weak, strong)
#no significant differences, suggests there is not measurement invariance and we may not be able to run stable/successful growth models.
```
Fit a second order growth model. Compare and contrast the estimates with the normal latent growth model.
```{r}
#Nevertheless, let's try to compute a second order growth model.
secondorder <- '
Dep1 =~ NA*T1BDI1 + L1*T1BDI1 + L2*T1BDI2 + L3*T1BDI4
Dep2 =~ NA*T2BDI1 + L1*T2BDI1 + L2*T2BDI2 + L3*T2BDI4
Dep3 =~ NA*T3BDI1 + L1*T3BDI1 + L2*T3BDI2 + L3*T3BDI4

T1BDI1 ~ t1*1
T1BDI2 ~ t2*1
T1BDI4 ~ t3*1

T2BDI1 ~ t1*1
T2BDI2 ~ t2*1
T2BDI4 ~ t3*1

T3BDI1 ~ t1*1
T3BDI2 ~ t2*1
T3BDI4 ~ t3*1

T1BDI1 ~~ T1BDI2 + T1BDI4
T1BDI2 ~~ T1BDI4
T2BDI1 ~~ T2BDI2 + T2BDI4
T2BDI2 ~~ T2BDI4
T3BDI1 ~~ T3BDI2 + T3BDI4
T3BDI2 ~~ T3BDI4

Dep1 ~ 0*1
Dep2 ~ 0*1
Dep3 ~ 0*1

L1 == 3 - L2 - L3
t1 == 0 - t2 - t3

i =~ 1*Dep1 + 1*Dep2 + 1*Dep3
s =~ 0*Dep1 + 1*Dep2 + 2*Dep3 '

secondorder<- growth(secondorder, data = DepressionSEMShort, missing = "ML")
summary(secondorder, fit.measures = TRUE)
```
Plot the second-order model
```{r}
semPaths(secondorder, whatLabels = "est", panelGroups = TRUE)
```
Simple Growth Model
```{r}
simplegrowth <- 'i_1 =~ 1*T1BDI1 + 1*T1BDI2 + 1*T1BDI4
                 s_1 =~ 0*T1BDI1 + 1*T1BDI2 + 2*T1BDI4

                 i_2 =~ 1*T2BDI1 + 1*T2BDI2 + 1*T2BDI4
                 s_2 =~ 0*T2BDI1 + 1*T2BDI2 + 2*T2BDI4

                 i_3 =~ 1*T3BDI1 + 1*T3BDI2 + 1*T3BDI4
                 s_3 =~ 0*T3BDI1 + 1*T3BDI2 + 2*T3BDI4
'

simplegrowth <- growth(simplegrowth, data = DepressionSEMShort)
summary(simplegrowth, fit.measures = TRUE)
#Looks like there are some negative covariances and a negative variance on the T3 slopes...
anova(secondorder, simplegrowth)
#Looks like the second order model fits better? Complicated by the warnings for the simplegrowth model though...
```

Plot the simple growth model
```{r}
semPaths(simplegrowth, whatLabels = "est", panelGroups = TRUE)
```
Fit a series of multiple group models. Constrain some parameters and compare the fit.

Let's try evaluating Affective sx of depression (BDI1, BDI2, BDI4) vs. Somatic sx. of deprsesion (BDI14, BDI15, BDI17) across group (gender)
```{r}
Group1 <- '
Aff1 =~ T1BDI1 + T1BDI2 + T1BDI4
Aff2 =~ T2BDI1 + T2BDI2 + T2BDI4
Aff3 =~ T3BDI1 + T3BDI2 + T3BDI4

Som1 =~ T1BDI14 + T1BDI15 + T1BDI17
Som2 =~ T2BDI14 + T2BDI15 + T2BDI17
Som3 =~ T3BDI14 + T3BDI15 + T3BDI17

Affective =~ Aff1 + Aff2 + Aff3
Somatic =~ Som1 + Som2 + Som3

Affective ~~ Somatic
'

fit.Group1 <- cfa(Group1, data = DepressionSEMShort, std.lv = TRUE, missing = "ML", group = "Gender")
summary(fit.Group1, standardized = TRUE, fit.measures = TRUE)
#Encountering some warnings with the cov and variances
```

Plotting:
```{r}
semPaths(fit.Group1, whatLabels = "est", panelGroups = TRUE)
```

Constraining parameters:

Constrain the Affective means first
```{r}
Group1.2 <- 'Aff1 =~ T1BDI1 + T1BDI2 + T1BDI4
Aff2 =~ T2BDI1 + T2BDI2 + T2BDI4
Aff3 =~ T3BDI1 + T3BDI2 + T3BDI4

Som1 =~ T1BDI14 + T1BDI15 + T1BDI17
Som2 =~ T2BDI14 + T2BDI15 + T2BDI17
Som3 =~ T3BDI14 + T3BDI15 + T3BDI17

Affective =~ c(L1, L1)*Aff1 + C(L2, L2)*Aff2 + c(L3, L3)*Aff3
Somatic =~ Som1 + Som2 + Som3

Affective ~~ Somatic
'

fit.Group1.2 <- cfa(Group1.2, data = DepressionSEMShort, group = "Gender")
summary(fit.Group1.2, standardized = TRUE, fit.measures = TRUE)
anova(fit.Group1, fit.Group1.2)
#Models are identical, constraining the affective means did not make the model worse, i.e., men and women don't differ on their affective symptoms of depression
```
Now try constraining the Somatic means:
```{r}
Group1.3 <- 'Aff1 =~ T1BDI1 + T1BDI2 + T1BDI4
Aff2 =~ T2BDI1 + T2BDI2 + T2BDI4
Aff3 =~ T3BDI1 + T3BDI2 + T3BDI4

Som1 =~ T1BDI14 + T1BDI15 + T1BDI17
Som2 =~ T2BDI14 + T2BDI15 + T2BDI17
Som3 =~ T3BDI14 + T3BDI15 + T3BDI17

Affective =~ Aff1 + Aff2 + Aff3
Somatic =~ c(L1, L1)*Som1 + C(L2, L2)*Som2 + c(L3, L3)*Som3

Affective ~~ Somatic
'

fit.Group1.3 <- cfa(Group1.3, data = DepressionSEMShort, group = "Gender")
summary(fit.Group1.3, standardized = TRUE, fit.measures = TRUE)
anova(fit.Group1, fit.Group1.3)
#Again, models are identical, constraining the somatic means also did not make the model worse, i.e., men and women don't differ on their somatic symptoms of depression
```
Try constraining the factor loadings:

```{r}
Group1.4 <- 'Aff1 =~ T1BDI1 + T1BDI2 + T1BDI4
Aff2 =~ T2BDI1 + T2BDI2 + T2BDI4
Aff3 =~ T3BDI1 + T3BDI2 + T3BDI4

Som1 =~ T1BDI14 + T1BDI15 + T1BDI17
Som2 =~ T2BDI14 + T2BDI15 + T2BDI17
Som3 =~ T3BDI14 + T3BDI15 + T3BDI17

Affective =~ Aff1 + Aff2 + Aff3
Somatic =~ Som1 + Som2 + Som3

Affective ~~ Somatic
'

fit.Group1.4 <- cfa(Group1.4, data = DepressionSEMShort, group = "Gender", group.equal = "loadings")
summary(fit.Group1.4, standardized = TRUE, fit.measures = TRUE)
anova(fit.Group1, fit.Group1.4)
#Still, models are identical, constraining the factor loadings suggests that men and women really don't differ that much on the constructs as the fit is similar, when loadings are constrained to be equal
```

Try constraining intercepts:
```{r}
Group1.5 <- 'Aff1 =~ T1BDI1 + T1BDI2 + T1BDI4
Aff2 =~ T2BDI1 + T2BDI2 + T2BDI4
Aff3 =~ T3BDI1 + T3BDI2 + T3BDI4

Som1 =~ T1BDI14 + T1BDI15 + T1BDI17
Som2 =~ T2BDI14 + T2BDI15 + T2BDI17
Som3 =~ T3BDI14 + T3BDI15 + T3BDI17

Affective =~ Aff1 + Aff2 + Aff3
Somatic =~ Som1 + Som2 + Som3

Affective ~~ Somatic
'

fit.Group1.5 <- cfa(Group1.5, data = DepressionSEMShort, group = "Gender", group.equal = "intercepts")
summary(fit.Group1.5, standardized = TRUE, fit.measures = TRUE)
anova(fit.Group1, fit.Group1.5)
#Still, models are identical, constraining the intercepts suggests that men and women really don't differ that much as the fit is similar, when intercepts are constrained to be equal
```


Finally, try constraining some, but not all parameterse:

```{r}
Group1.6 <- 'Aff1 =~ T1BDI1 + T1BDI2 + T1BDI4
Aff2 =~ T2BDI1 + T2BDI2 + T2BDI4
Aff3 =~ T3BDI1 + T3BDI2 + T3BDI4

Som1 =~ T1BDI14 + T1BDI15 + T1BDI17
Som2 =~ T2BDI14 + T2BDI15 + T2BDI17
Som3 =~ T3BDI14 + T3BDI15 + T3BDI17

Affective =~ Aff1 + Aff2 + Aff3
Somatic =~ Som1 + Som2 + Som3

Affective ~~ Somatic
'

fit.Group1.6 <- cfa(Group1.6, data = DepressionSEMShort, group = "Gender", group.equal = c("loadings", "intercepts"), group.partial = c("Affective =~ Aff1 + Aff2 + Aff3", "Somatic =~ Som1 + Som2 + Som3"))
summary(fit.Group1.6, standardized = TRUE, fit.measures = TRUE)
anova(fit.Group1, fit.Group1.6)
#Aaaand the models are stil identical, constraining the loadings and intercepts doesn't change fit, again suggesting that men and women really don't differ that much on experience of depression
```



