---
title: "Growth Curves Week 2"
author: "Marilyn Piccirillo"
date: "September 10, 2017"
output: pdf_document
toc = yes
---
#Data Prep
```{r}
rm(list = ls())
library(foreign)
library(tidyverse)
library(plyr)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(lubridate)
library(lme4)

#Pull in all three timepoints
SuicideT1 <- read.spss("StressStudy_T1short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT2 <- read.spss("StressStudy_T2short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT3 <- read.spss("StressStudy_T3short.sav", use.value.labels = FALSE, to.data.frame = TRUE)

#Merge all three timepoints
SuicideMerge <- left_join(SuicideT1, SuicideT2, by = "ID")
SuicideMerge <- left_join(SuicideMerge, SuicideT3, by= "ID")

#Separatedates
SuicideMerge$T1_Date <- as.Date(SuicideMerge$T1_DateStarted)
SuicideMerge$T2_Date <- as.Date(SuicideMerge$T2_DateStarted)
SuicideMerge$T3_Date <- as.Date(SuicideMerge$T3_DateStarted)

#Make a shorter dataset
Suicideshort <- dplyr::select(SuicideMerge, ID, T1_Date, T2_Date, T3_Date, T1_BSSTot, T2_BSSTot, T3_BSSTot, T1_BDITot, T2_BDITot, T3_BDITot, T1_SIASTot, T2_SIASTot, T3_SIASTot, T1_ACSSTot, T2_ACSSTot, T3_ACSSTot, Gender, Age, Ethn, SexOrien)
#Converting into proper format
Suicideshort$Gender <- as.factor(Suicideshort$Gender)
Suicideshort$Ethn <- as.factor(Suicideshort$Ethn)
Suicideshort$SexOrien <- as.factor(Suicideshort$SexOrien)
Suicideshort$Age <- as.character(Suicideshort$Age)
Suicideshort$Age <- as.numeric(Suicideshort$Age)
Suicideshort <- as.tibble(Suicideshort)

#Convert SuicideShort to longform
Suicidelong <- gather(Suicideshort, Var, Val, select = c("T1_ACSSTot", "T1_BDITot", "T1_BSSTot", "T1_SIASTot", "T2_ACSSTot", "T2_BDITot", "T2_BSSTot", "T2_SIASTot", "T3_ACSSTot", "T3_BDITot", "T3_BSSTot", "T3_SIASTot"))
Suicidelong <- arrange(Suicidelong, .by_group = ID)
Suicidelong <- separate (Suicidelong, Var, c("Timepoint", "Variable"), sep = "_")
Suicidelong <- spread(Suicidelong, Variable, Val)
Suicidelong <- as.tibble(Suicidelong) 

rm(SuicideT1, SuicideT2, SuicideT3, SuicideMerge)

Suicidelong$WhichDate <- ifelse(!is.na (Suicidelong$T3_Date), "No", "Yes") #Create a variable to use for ifelse
Suicidelong$Days <- ifelse(Suicidelong$WhichDate=="Yes", (interval(Suicidelong$T1_Date, Suicidelong$T2_Date)), (interval(Suicidelong$T1_Date, Suicidelong$T3_Date))) #Instructs R to use T3 if avaiable or T2 if not available
Suicidelong$Days <- as.duration(Suicidelong$Days)
Suicidelong$Days <- as.numeric(Suicidelong$Days) / (3600*24) #Converting Days into a readable date
Suicidelong$WhichDate <- NULL #Removes WhichDate variable

#DECIDING TO ONLY KEEP PPL WHO ONLY COMPLETED THREE TIME POINTS
#Remove participants who only completed *ONE OR TWO* timepoints
Suicide2 <- tally(group_by(Suicidelong, ID, BSSTot), sort = TRUE)
tally(Suicide2, sort = TRUE)
tally(tally(Suicide2))
Suicide2 <- arrange(Suicide2, ID)
Suicide2 <- Suicide2 %>%
  filter(!is.na(BSSTot)) #This created a separate dataset that tallied all "real" timepoints, based on BSS
Suicide2$BSSTot<- NULL
#Merge n variable into Suicide long
Suicidelong <- right_join(Suicide2, Suicidelong, by = "ID") #Now I can filter Suicidelong based on n
#Remove all rows from Suicidelong that only have 1 value
Suicidelong <- filter(Suicidelong, n > 2)
#Now Suicidelong only has data from people who completed >1 timepoint.

rm(Suicideshort, Suicide2)
```

#Running a linear model
##1.Run linear models on all of your subjects (a basic regression). What is the average intercept, the average slope?
```{r}
Dep1 <- lm(BDITot ~ Days, data = Suicidelong)
summary(Dep1)
#Average intercept = 6.746. The mean BDI score at Time 1 is 6.746 (minimal depression)
#Average slope = -0.003. The average amount of change over time is -0.003 (slightly decreasing depression over time)
```

#Running a MLM/LMER model
##Now run a mlm/lmer model with only a random intercept. What is the ICC? What does residual variance look like compared to linear model? Create a graph to show this effect.
```{r}
#Run a MLM/LMER model with only a random intercept
library(lme4)
Dep2 <- lmer(BDITot ~ 1 + (1 | ID), data = Suicidelong)
summary(Dep2)

#What is the ICC?
library(sjstats)
icc(Dep2)
#ICC = .612. It looks like about 60% of the variance is attributable to between-person variation. 

#What does residual variance look like compared to linear model? Create a graph to show this effect
library(broom)
library(ggplot2)
library(ggthemes)
#Extract the model information
Dep2_fitted = augment(Dep2, Suicidelong)
Dep1_fitted = augment(Dep1, Suicidelong)

#Create a new matrix to store the residual information
Dep_residual_plot = matrix(nrow = 2, ncol = 2) #Setting up matrix
colnames(Dep_residual_plot) = c("Model", "SE_residuals") #Naming columns
Dep_residual_plot = as.data.frame(Dep_residual_plot) #Making matrix into a dataframe
Dep_residual_plot$Model = c("Linear Model", "MLM (no intercept)") #Pulling in the information
Dep_residual_plot$SE_residuals = c(sd(Dep1_fitted$.resid), sd(Dep2_fitted$.resid)) #Converting std. error into std. dev?

#Plotting the residuals
ggplot(Dep_residual_plot, aes(x = Model, y = SE_residuals)) + 
  geom_bar(fill = "salmon2", stat = "identity", width = 0.5) +
  theme_few()+
  ggtitle("Standard Error of Residuals \n for Linear Model & MLM")

#Looks like there is more residual SE in the linear model as compared to the intercept-only model, which means that the intercept-only model (MLM) does a better job of modeling depression over time.
```

#Running a MLM model with a fixed slopes term
##Introduce a fixed slope term. What is the difference in terms of the fixed effects estimates between this estimate and the previous? Of the residual standard error? Create a graph to show both fixed effects estimates and the CIs around them.
```{r}
#Introduce a fixed slope term. What is the difference in terms of the fixed effects esimates between this estimate and the previous?
Dep3 <- lmer(BDITot ~ Days + (1 | ID), data = Suicidelong)
summary(Dep3)
summary(Dep2)
#This model includes a fixed effect of Time. How do individual's suicide scores change over time?
#With the fixed effect estimate of the change in suicide score as a function of time, we now can add a slope to our regression lines for each individual. The intercept only model fit parallel lines with slope 0 for each subject. Now we can fit parallel lines for each subject (effect is fixed), but with slope = -0.003.

#What is the difference in terms of the residual standard error between the two models?
RSE_Dep2 <- rse(Dep2)
RSE_Dep3 <- rse(Dep3)
RSE_Dep2
RSE_Dep3
#RSE has increased slightly in the model with the fixed effect as compared to intercept only model. This suggests that the fixed effect model performs slightly worse than the intercept only model in terms of modeling depression over time.
RSEdiff <- RSE_Dep3-RSE_Dep2
RSEdiff
#However the fixed-effect model only explains .004 less residual variance as compared to the intercept-only model. This means they are basically equally good at describing variance...maybe this means that depression scores didn't really change much over time? That if you score high at T1, you'll also score high throughout the rest of the study?

#Create a graph to show both fixed effects estimates and the CIs around them
GraphDepFE <- sjp.lmer(Dep3, type = "fe")
GraphDepFE
#The confidence interval looks really large though...does this mean there is a lot of individual variability, despite not being statistically significant?
```

#Running a MLM model with a random slope term
##Run an additional model with a random slope. How does this change compare to the previous model? Should you keep the random slope or not?
```{r}
Dep4 <- lmer(BDITot ~ Days + (Days|ID), data = Suicidelong)
summary(Dep4)

RSE_Dep4 <- rse(Dep4)
RSE_Dep4

RSEdiff2 <- RSE_Dep3-RSE_Dep4
RSEdiff2

#Model with the random effects has more residual variance so it may be worse at explaining the variance in the model? Can also compare the two models using anova

anova(Dep3, Dep4)
#The test is not significant suggesting that the random slopes model does not explain significnatly less variance and thus, it is not a worse model. To save on computational power (and since the random slopes aren't that useful), maybe we should use the fixed effects only model?
```

##Interpret the correlation between the slope and the intercept.
```{r}
summary(Dep4)
#The correlation of fixed effects is -.947 which means the slope and intercept are highly negatively correlated. Thus, if someone starts out with a high intercept, they will have a very small slope or vice versa?
```

##Create a density plot of the random effects from your final model.
```{r}
library(merTools)
Dep_re.sim = REsim(Dep4)

Dep_fe.sim = FEsim(Dep4)

library(ggplot2)
Graph_Dep4 <- Dep_re.sim %>%
  filter(term == "(Intercept)")

ggplot(Graph_Dep4, aes(mean)) +
  geom_density() +
    theme_few()+
  ggtitle("Dep: Density Plot for Intercept")

Graph_Dep4a <- Dep_re.sim %>%
  filter(term == "Days")

ggplot(Graph_Dep4a, aes(mean)) +
  geom_density() +
    theme_few()+
  ggtitle("Dep: Density Plot for Days")
```

##Create a catepilar plot of the random effects. Is there any person that seems odd in terms of a large standard errors around intercept and slope estimates?
```{r}
Dep_p1 = plotREsim(Dep_re.sim)
Dep_p1
#There does seem to be one person who has a large intercept (indicating they are really depressed at Time1)
```

##Create a plot of the trajectory, along with a spaghetti plot of each person's individual slope. Set the alpha level (transparency) on the individual slopes to make them easier to see.

```{r}
tbl_df(Suicidelong) %>% 
  mutate(pred = predict(Dep4, newdata = .)) %>%
  ggplot(aes(x = Days, y = pred, group = "ID", color = ID)) + 
  geom_line(size = .5, alpha = .5) +
  labs(x = "Days", y = "Predicted Depression Scores", 
       title = "Predicted Depression Scores From The Random Slope Model") + 
  theme_classic() + 
  theme(axis.text = element_text(face = "bold", size = rel(1.2)),
        axis.title = element_text(face = "bold", size = rel(1.2)),
        legend.title = element_text(face = "bold", size = rel(1.2)),
        plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```
        