---
title: "Making a Table"
author: "Marilyn Piccirillo"
date: "October 9, 2017"
output: pdf_document
---
Below is code that *should* work for all models. Just run the function and save it as an `R` object. You can use this with `papaja` and the `apa_table()` function pretty easily. The trick is that if you are not using the papaja template, the proper LaTeX packages may not be loaded. You can get around this by attaching a .tex file calling the packages under "in_header: header.tex" in your YAML header. The YAML header of this .Rmd file contains the necessary syntax and the header.tex file with the proper packages.  
```{r, echo = FALSE}
rm(list = ls())
library(foreign)
library(tidyverse)
library(plyr)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(lubridate)
library(lme4)
library(psych)
library(sjPlot)
library(MuMIn)
library(merTools)
library(reghelper)
library(stargazer)
library(lsmeans)
library(multcompView)

#Pull in all three timepoints
SuicideT1 <- read.spss("StressStudy_T1short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT2 <- read.spss("StressStudy_T2short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT3 <- read.spss("StressStudy_T3short.sav", use.value.labels = FALSE, to.data.frame = TRUE)

#Merge all three timepoints
SuicideMerge <- left_join(SuicideT1, SuicideT2, by = "ID")
SuicideMerge <- left_join(SuicideMerge, SuicideT3, by= "ID")

#Separatedates
SuicideMerge$T1_Date <- as.Date(SuicideMerge$T1_DateStarted)
SuicideMerge$T2_Date <- as.Date(SuicideMerge$T2_DateStarted)
SuicideMerge$T3_Date <- as.Date(SuicideMerge$T3_DateStarted)

#Make a shorter dataset
Suicideshort <- dplyr::select(SuicideMerge, ID, T1_Date, T2_Date, T3_Date, T1_BSSTot, T2_BSSTot, T3_BSSTot, T1_BDITot, T2_BDITot, T3_BDITot, T1_SIASTot, T2_SIASTot, T3_SIASTot, T1_ACSSTot, T2_ACSSTot, T3_ACSSTot, Gender, Age, Ethn, SexOrien)
#Converting into proper format
Suicideshort$Gender <- as.factor(Suicideshort$Gender)
Suicideshort$Ethn <- as.factor(Suicideshort$Ethn)
Suicideshort$SexOrien <- as.factor(Suicideshort$SexOrien)
Suicideshort$Age <- as.character(Suicideshort$Age)
Suicideshort$Age <- as.numeric(Suicideshort$Age)
Suicideshort <- as.tibble(Suicideshort)

#Convert SuicideShort to longform
Suicidelong <- gather(Suicideshort, Var, Val, select = c("T1_ACSSTot", "T1_BDITot", "T1_BSSTot", "T1_SIASTot", "T2_ACSSTot", "T2_BDITot", "T2_BSSTot", "T2_SIASTot", "T3_ACSSTot", "T3_BDITot", "T3_BSSTot", "T3_SIASTot"))
Suicidelong <- arrange(Suicidelong, .by_group = ID)
Suicidelong <- separate (Suicidelong, Var, c("Timepoint", "Variable"), sep = "_")
Suicidelong <- spread(Suicidelong, Variable, Val)
Suicidelong <- as.tibble(Suicidelong) 

rm(SuicideT1, SuicideT2, SuicideT3, SuicideMerge)

Suicidelong$WhichDate <- ifelse(!is.na (Suicidelong$T3_Date), "No", "Yes") #Create a variable to use for ifelse
Suicidelong$Days <- ifelse(Suicidelong$WhichDate=="Yes", (interval(Suicidelong$T1_Date, Suicidelong$T2_Date)), (interval(Suicidelong$T1_Date, Suicidelong$T3_Date))) #Instructs R to use T3 if avaiable or T2 if not available
Suicidelong$Days <- as.duration(Suicidelong$Days)
Suicidelong$Days <- as.numeric(Suicidelong$Days) / (3600*24) #Converting Days into a readable date
Suicidelong$WhichDate <- NULL #Removes WhichDate variable

#DECIDING TO ONLY KEEP PPL WHO ONLY COMPLETED THREE TIME POINTS
#Remove participants who only completed *ONE OR TWO* timepoints
Suicide2 <- tally(group_by(Suicidelong, ID, BSSTot), sort = TRUE)
tally(Suicide2, sort = TRUE)
tally(tally(Suicide2))
Suicide2 <- arrange(Suicide2, ID)
Suicide2 <- Suicide2 %>%
  filter(!is.na(BSSTot)) #This created a separate dataset that tallied all "real" timepoints, based on BSS
Suicide2$BSSTot<- NULL
#Merge n variable into Suicide long
Suicidelong <- right_join(Suicide2, Suicidelong, by = "ID") #Now I can filter Suicidelong based on n
#Remove all rows from Suicidelong that only have 1 value
Suicidelong <- filter(Suicidelong, n > 2)
#Now Suicidelong only has data from people who completed >1 timepoint.

rm(Suicideshort, Suicide2)
Dep4 <- lmer(BDITot ~ Days + (Days|ID), data = Suicidelong)
```

```{r, echo = FALSE}
## here's some code to make a table. You shouldn't need to modify anything here 
# unless you add additional random effects terms
## fixed effects first ##
table_fun <- function(model){
  fixed <- broom::tidy(model) %>% 
    filter(group == "fixed") %>%
    dplyr::select(term, estimate) 
  ## add random effects ##
  rand <- broom::tidy(model) %>% 
    filter(group != "fixed") %>%
    dplyr::select(term, estimate)
  ## get confidence intervals ##
  CI <- data.frame(confint.merMod(model, method = "boot", nsim = 10, boot.type = "perc"))%>%
    mutate(term = rownames(.)) %>% 
    setNames(c("lower", "upper", "term"))
  
  ## Get ICC & R2 values ##
  ICC <- reghelper::ICC(model)
  R2 <- MuMIn::r.squaredGLMM(model)
  
  ## format the fixed effects
  fixed <- fixed %>% left_join(CI %>% filter(!grepl(".sig", term))) %>%
    mutate(type = "Fixed Parts")
  
  rand <- rand %>%
    mutate(estimate = ifelse(grepl("cor", term) == T, estimate, estimate^2),
          term = mapvalues(term, unique(term), 
           c("$\\tau_{00}$", "$\\tau_{11}$", "$\\tau_{10}$", "$\\hat{\\sigma^2}$"))) %>%
    left_join(
      CI %>% filter(grepl(".sig", term)) %>%
        mutate(term = mapvalues(term, unique(term), 
            c("$\\tau_{00}$", "$\\tau_{10}$", "$\\tau_{11}$", "$\\hat{\\sigma^2}$")),
            lower = lower^2, upper = upper^2)) %>%
    mutate(type = "Random Parts")
  
  mod_terms <- tribble(
    ~term, ~estimate, ~type,
    "ICC", ICC, "Model Terms",
    "$R^2_m$", R2[1], "Model Terms",
    "$R^2_c$", R2[2], "Model Terms"
  )
  
  tab <- fixed %>%
    full_join(rand) %>%
    mutate(CI = sprintf("(%.2f, %.2f)", lower, upper)) %>%
    dplyr::select(-lower, -upper) %>%
    full_join(mod_terms) %>%
    mutate(estimate = sprintf("%.2f", estimate)) %>%
    dplyr::rename(b = estimate) %>%
    dplyr::select(type, everything())
  return(tab)
}
tab <- table_fun(Dep4)
# you can use this with papaja and the apa_table function pretty easily
# the trick is that if you are not using the papaja template, the proper
# LaTeX packages may not be loaded. You can get around this by attaching
# a .tex file calling the packages under "in_header: header.tex" in your YAML
# header the YAML header of this .Rmd file contains the necessary syntax and 
# the header.tex file with the proper packages
options(knitr.kable.NA = '')
knitr::kable(tab, caption = "Ugly MLM Table Example")
```


### More Advanced: `kable()` + `kableExtra`  
\small
```{r, results = 'asis'}
library(kableExtra)
options(knitr.kable.NA = '')
knitr::kable(tab %>% #select(-type) %>%
    mutate(term = gsub("[()]", "", term)),
             caption = "Not Quite Right kableExtra MLM Table Example", 
    format = "latex", 
    #longtable = T, 
    booktabs = T, escape = F) %>%
  # group_rows("Fixed", 1,4) %>% 
  # group_rows("Random", 5,9) %>%
  # group_rows("Model", 9,11) %>%
  collapse_rows(1) %>%
  #kable_styling(latex_options = c("striped","repeat_header"),full_width = F)
  add_header_above(c(" ", " ", "Model 1" = 2))
```


### Alternative: `papaja` + `apa_table()`  
```{r, results = 'asis'}
papaja::apa_table(tab %>% dplyr::select(-type),caption = "papaja MLM Table Example", 
    na_string = "", stub_indents = list(Fixed = c(1:4), Random = c(5:8), Summary = c(9:11)),
    col_spanners = list(`Depression scores` = c(2,3)))
```