---
title: "ALDA_SEM"
author: "Marilyn Piccirillo"
date: "10/19/2017"
output: pdf_document
---
Plans:
1) Present the MLM models
	one with intercept only
	one with a fixed slope
	one with a random slope
	include some predictors

2) Present the SEM models
	one with intercept only
	one with a fixed slope
	one with a random slope
  talk about constraining the slope to have no variance in SEM (e.g., y1 ~~ 0* y5, when y5 =   slope)
  include some predictors

3) Look at the estimators
4) Discuss missing data
5) How to get bootstrapped estimate
6) How to compare models (using anova function)
7) Discuss centering, chaning the time metric.
############################################################################################

Data Prep:
Undergraduate students completed self-report measures (and some clinical measures) three times, 6 weeks apart. Unfortunately, there was lots of attrition...over 500 at the first time point, approx. 150 at the third time point. Only ~100 people completed all three timepoints. For the sake of the class, I'm going to only show models using people who completed all three time points, although this introduces a lot of sample bias :o

I was originally interested in modeling suicidality which was measured in this study through the Beck Suicide Scale, which asks questions about whether or not they've ever thought about suicide, had a plan, made an attempt, etc. Although a surprisingly high number of people do have  scores for this measure, it's not really enough to model with. I instead chose to model depression scores over the 3- month period using the BDI-II.
```{r}
rm(list = ls())
library(foreign)
library(tidyverse)
library(plyr)
library(dplyr)
library(tidyr)
library(broom)
library(ggplot2)
library(lubridate)
library(lme4)
library(lavaan)
library(sjstats)
library(sjPlot)

#Pull in all three timepoints
SuicideT1 <- read.spss("StressStudy_T1short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT2 <- read.spss("StressStudy_T2short.sav", use.value.labels = FALSE, to.data.frame = TRUE)
SuicideT3 <- read.spss("StressStudy_T3short.sav", use.value.labels = FALSE, to.data.frame = TRUE)

#Merge all three timepoints
SuicideMerge <- left_join(SuicideT1, SuicideT2, by = "ID")
SuicideMerge <- left_join(SuicideMerge, SuicideT3, by= "ID")

#Separatedates
SuicideMerge$T1_Date <- as.Date(SuicideMerge$T1_DateStarted)
SuicideMerge$T2_Date <- as.Date(SuicideMerge$T2_DateStarted)
SuicideMerge$T3_Date <- as.Date(SuicideMerge$T3_DateStarted)

#Make a shorter dataset
Suicideshort <- dplyr::select(SuicideMerge, ID, T1_Date, T2_Date, T3_Date, T1_BSSTot, T2_BSSTot, T3_BSSTot, T1_BDITot, T2_BDITot, T3_BDITot, T1_SIASTot, T2_SIASTot, T3_SIASTot, T1_ACSSTot, T2_ACSSTot, T3_ACSSTot, Gender, Age, Ethn, SexOrien)
#Converting into proper format
Suicideshort$Gender <- as.factor(Suicideshort$Gender)
Suicideshort$Ethn <- as.factor(Suicideshort$Ethn)
Suicideshort$SexOrien <- as.factor(Suicideshort$SexOrien)
Suicideshort$Age <- as.character(Suicideshort$Age)
Suicideshort$Age <- as.numeric(Suicideshort$Age)
Suicideshort <- as.tibble(Suicideshort)

#Make date variables
Suicideshort$T1_Day <- difftime(Suicideshort$T1_Date, Suicideshort$T1_Date, units = c("days"))
Suicideshort$T2_Day <- difftime(Suicideshort$T2_Date, Suicideshort$T1_Date, units = c("days"))
Suicideshort$T3_Day <- difftime(Suicideshort$T3_Date, Suicideshort$T1_Date, units = c("days"))

#Selecting only people who completed all three time points
Suicideshort <- subset(Suicideshort, T1_BSSTot >= "0" & T2_BSSTot >= "0" & T3_BSSTot >= "0")

#Convert SuicideShort to longform
Suicidelong <- gather(Suicideshort, Var, Val, select = c("T1_ACSSTot", "T1_BDITot", "T1_BSSTot", "T1_SIASTot", "T2_ACSSTot", "T2_BDITot", "T2_BSSTot", "T2_SIASTot", "T3_ACSSTot", "T3_BDITot", "T3_BSSTot", "T3_SIASTot", "T1_Day", "T2_Day", "T3_Day"))
Suicidelong <- arrange(Suicidelong, .by_group = ID)
Suicidelong <- separate (Suicidelong, Var, c("Timepoint", "Variable"), sep = "_")
Suicidelong <- spread(Suicidelong, Variable, Val)
Suicidelong <- as.tibble(Suicidelong) 
#Removing someone with bad T2 data
Suicidelong <- Suicidelong[-c(784:786), ]

Suicidelong <- dplyr::select(Suicidelong, -Timepoint)

#Create a wave variable...repeat 1,2,3 for each person
n<- c(1, 2, 3)
Suicidelong$Wave <- rep(n, 130)
#Make a shorter dataset to work with.
Suicidelong <- dplyr::select(Suicidelong, ID, Wave, Day, Gender, Age, Ethn, SexOrien, BDITot, BSSTot, SIASTot, ACSSTot)

#Make Wave variables
Suicideshort$Wave1 <- rep(1, 130)
Suicideshort$Wave2 <- rep(2, 130)
Suicideshort$Wave3 <- rep(3, 130)
#Shorten and rearrange Suicideshort
Suicideshort <- dplyr::select(Suicideshort, ID, Wave1, Wave2, Wave3, T1_Day, T2_Day, T3_Day, Gender, Age, Ethn, SexOrien, T1_BDITot, T2_BDITot, T3_BDITot, T1_BSSTot, T2_BSSTot, T3_BSSTot, T1_SIASTot, T2_SIASTot, T3_SIASTot, T1_ACSSTot, T2_ACSSTot, T3_ACSSTot)

#Clean up
rm(SuicideT1, SuicideT2, SuicideT3, SuicideMerge)
```

1) Visualize the data
Let's first show the study variables with the entire dataset. Need to finish up the full dataset.
```{r}
# #Suicidelong %>%
#   #ggplot(aes(x = Day, y = BDITot)) + 
#     geom_smooth(aes(group = ID), method = "lm", se = F, color = "gray", size = .5) +
#     geom_smooth(method = "lm", se = F, color = "red", size = 1) +
#     labs(x = "Days", y = "Depression Score", title = "Simple Growth Curve") +
#     theme_classic() +
#     theme(legend.position = "none",
#           axis.text = element_text(face = "bold", size = rel(1.2)),
#           axis.title = element_text(face = "bold", size = rel(1.2)),
#           plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
# 
# boxplot(Suicidelongfull$BDITot,
#         main = "Boxplot of Depression scores",
#         xlab = "Depression",
#         ylab = "Score",
#         ylim = c(0, 65))
# boxplot(Suicidelongfull$ACSSTot,
#         main = "Boxplot of Acquired Capability for Suicide scores",
#         xlab = "Fear of Death",
#         ylab = "Score",
#         ylim = c(0, 20))
# boxplot(Suicidelongfull$SIASTot,
#         main = "Boxplot of Social Anxiety scores",
#         xlab = "Social Anxiety",
#         ylab = "Score",
#         ylim = c(0, 70))
# boxplot(Suicidelongfull$BSSTot,
#         main = "Boxplot of Suicide scores",
#         xlab = "Suicide",
#         ylab = "Score",
#         ylim = c(0, 30))
```
#We see there's a pretty good range of scores. For reference: 19% have #moderate depression, 8% have severe depression. 31% have suggested #clincial levels of social interaction anxiety.


###############START HERE#########################
Now lets look at the descriptives for the subset of those who completed all three timepoints
```{r}
boxplot(Suicidelong$BDITot,
        main = "Boxplot of Depression scores",
        xlab = "Depression",
        ylab = "Score",
        ylim = c(0, 65))
boxplot(Suicidelong$ACSSTot,
        main = "Boxplot of Acquired Capability for Suicide scores",
        xlab = "Fear of Death",
        ylab = "Score",
        ylim = c(0, 20))
boxplot(Suicidelong$SIASTot,
        main = "Boxplot of Social Anxiety scores",
        xlab = "Social Anxiety",
        ylab = "Score",
        ylim = c(0, 70))
boxplot(Suicidelong$BSSTot,
        main = "Boxplot of Suicide scores",
        xlab = "Suicide",
        ylab = "Score",
        ylim = c(0, 30))
```


```{r}
Suicidelong %>%
  ggplot(aes(x = Day, y = BDITot)) + 
    geom_smooth(aes(group = ID), method = "lm", se = F, color = "gray", size = .5) +
    geom_smooth(method = "lm", se = F, color = "red", size = 1) +
    labs(x = "Days", y = "Depression Score", title = "Simple Growth Curve") +
    theme_classic() +
    theme(legend.position = "none",
          axis.text = element_text(face = "bold", size = rel(1.2)),
          axis.title = element_text(face = "bold", size = rel(1.2)),
          plot.title = element_text(face = "bold", size = rel(1.2), hjust = .5))
```
We see some variability here in how people's scores change over the number of days in the study. 


2) Run MLM models:
Running an empty (intercept only) model.
```{r}
IntOnly <- lmer(BDITot ~ 1 + (1 | ID), data = Suicidelong)
summary(IntOnly)
icc(IntOnly)
```
Looks like there's roughly an equal distribution between the within and between person levels (slightly more on the between-person level). Also, people significantly vary in terms of where they start out in terms of depression.

Running a fixed slopes model.
```{r}
FixSlopes <- lmer(BDITot ~ Day + (1 | ID), data = Suicidelong)
summary(FixSlopes)
```
It looks like intercept and slope are moderately correlated. That is, we see that people start out at different levels of depression (significant, positive intercept), and they tend to decrease over time (significant, negative slope). These could be semester effects, but this study ran over the entire year, so it's likely these seasonal effects were averaged out. 

Let's see how the fixed slope model compares to the intercept-only model
```{r}
rse(IntOnly)
rse(FixSlopes)

anova(IntOnly, FixSlopes)
```
It looks like adding in the fixed slopes helps to account for more of the residual standard error and is a signficantly better fitting model than the intercept only model. 


Now let's try to model a random slope
```{r}
RandSlope <- lmer(BDITot ~ Day + (Day|ID), data = Suicidelong)
summary(RandSlope)
```
With the random slope model, we see a stronger correlation between intercept and slope. This means that the higher you start out in depression, the more you are expected to decrease over time (i.e., regression to the mean). We also see in this model that the intercept is still positive and statistically significant. When we model a random slope that accounts for individual variability, we see that people on average tend to decrease in their level of depression and this is statistically significant.

Checking rse again. Also let's compare all the models.
```{r}
rse(IntOnly)
rse(FixSlopes)
rse(RandSlope)

anova(IntOnly, FixSlopes)
anova(IntOnly, RandSlope)
anova(FixSlopes, RandSlope)
```

Adding in the random slope helps account for more of the variance and it is significantly better than both the intercept-only model and the fixed slopes model.

 
Let's add in a predictors to see if things change...

We could see if the model changes based on the individual's socail anxiety scores. Note, social anxiety was assessed at all three time points, so this is technically a time-varying predictor
```{r}
SIAS <- lmer(BDITot ~ Day + SIASTot + Day*SIASTot + (Day|ID), data = Suicidelong)
summary(SIAS)
```
We see that all terms are statistically significant. That is, people who are more socially anxious are also more likely to be depressed. We also see that these are the people who are likely to decrease in their depression scores the most over time?

Let's try and graph this:
```{r}
sjp.int(SIAS, swap.pred = TRUE, mdrt.values = c("meansd"), xlim = c(0, 110), ylim = c(0, 20))
```
Let's try another predictor, fearlessness towards death (acquired capability for suicide)

```{r}
ACSS <- lmer(BDITot ~ Day + ACSSTot + Day*ACSSTot + (Day|ID), data = Suicidelong)
summary(ACSS)
sjp.int(ACSS, swap.pred = TRUE, mdrt.values = c("meansd"), xlim = c(0, 120), ylim = c(0, 20))
```
We see that people who start out with higher levels of fearlessness towards death are also more likely to be depressed, but there is no significant interaction with time.

Let's try and see if changing the time variable changes things. I've been measuring time in days since starting the study. What if we looked at these models using a wave variable.

```{r}
WFixSlopes <- lmer(BDITot ~ 1 + Wave + (1 | ID), data = Suicidelong)
summary(WFixSlopes)
```
Here we see that there is still a significant decrease in depression over time, but that the effect is stronger when we use the Wave variable instead of the Day variable

With a random slope:
```{r}
WRandSlope <- lmer(BDITot ~ Wave + (Wave|ID), data = Suicidelong)
summary(WRandSlope)
```
Similar. There is still a significant decrease in the effect of time, but the effect is stronger when using the Wave variable.

With a continuous, time-varying predictor:
```{r}
WSIAS <- lmer(BDITot ~ Wave + SIASTot + Wave*SIASTot + (Wave|ID), data = Suicidelong)
summary(WSIAS)
```
When using the Wave variable, the interaction with time is no longer significant.

Results plotted:
```{r}
sjp.int(WSIAS, swap.pred = TRUE, mdrt.values = c("meansd"), ylim = c(0, 20))
```


####################################################
Now let's try in SEM!

First we need to put data back into wide format:
```{r}
#FILL IN CODE HERE.
```

Let's first start with an intercept only model:
```{r}
SEMIntOnly <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot '
SEMIntOnlyfit <- growth(SEMIntOnly, orthogonal = TRUE, missing = "ML", data = Suicideshort)
summary(SEMIntOnlyfit)

SEMIntOnly2 <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot '
SEMIntOnly2fit <- sem(SEMIntOnly2, orthogonal = TRUE, missing = "ML", data = Suicideshort)
summary(SEMIntOnly2fit)
```
Looks like the intercept is similar to the MLM version (7.267 vs. 8.008)
When I compare it to a SEM model, the estimate for intercept = 0.
This model suggests that people have different average BDI scores across all time points; whereas the intercept MLM model suggests that people have different starting BDI scores.


Now let's try with a fixed slope:
```{r}
SEMFixSlopes_cov <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot

                  s ~~ 0*s #fixed slopes, no variance'
SEMFixSlopescovfit <- growth(SEMFixSlopes_cov, missing = "ML", data = Suicideshort)
summary(SEMFixSlopescovfit)
```

**Get an error for negative cov between slope and intercept**

When we compare these effects to the MLM models using the Wave variable, we see similar effects for the Slope and Intercept

```{r}
SEMFixSlopes <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot
                  i ~~ 0*s #no covariance 
                  s ~~ 0*s #fixed slopes, no variance'
SEMFixSlopesfit <- growth(SEMFixSlopes, missing = "ML", data = Suicideshort)
summary(SEMFixSlopesfit)
                  
```


Trying with a random slope:
```{r}
SEMRandSlopes <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot'
SEMRandSlopesfit <- growth(SEMRandSlopes, orthogonal = TRUE, missing = "ML", data = Suicideshort)
summary(SEMRandSlopesfit)
inspect(SEMRandSlopesfit, "cov.lv")
```
**Errors for negative variances for slope**

Again, when we compare the MLM Random Slope model, we see the same effects for slope and intercept.

Practice constraining the slope to have no variance
```{r}
SEMConstrainSlope <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot
                  s~~0*s'
SEMConstrainSlopefit <- growth(SEMConstrainSlope, orthogonal = TRUE, missing = "ML", data = Suicideshort)
summary(SEMConstrainSlopefit)
inspect(SEMConstrainSlopefit, "cov.lv")
```

Comparing fit from MLM versus SEM models:
```{r}
anova(SEMIntOnlyfit, IntOnly)
anova(SEMFixSlopesfit, WFixSlopes)
anova(SEMRandSlopesfit, WRandSlope)
```

Looks like there are significant differences between the MLM and SEM models...

Let's add in a predictor to the SEM model:
```{r}
SEM_SIAS <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot
                  s~~0*s

                  i ~ T1_SIASTot + T2_SIASTot + T3_SIASTot
                  s ~ T1_SIASTot + T2_SIASTot + T3_SIASTot'
SEM_SIASfit <- growth(SEM_SIAS, orthogonal = TRUE, missing = "ML", data = Suicideshort)
summary(SEM_SIASfit)
inspect(SEM_SIASfit, "cov.lv")
```
It looks like that people with higher SIAS scores at T2 also have significantly higher average BDI scores across time. People with higher SIAS scores at T1 show a significant reduction in BDI scores from T1 to T2, but people with higher SIAS scores at T3 demonstrate a signficant increase in BDI scores from T2 to T3.

*Why were there no errors for this model, even though there is a negative variance between slope and intercept?

########ESTIMATORS#########
Now let's see what happens with a different estimator. We have been using the ML estimator; however, this is best when the variables are multivariate normal. Let's see if BDI scores are really normal...
```{r}
hist(Suicideshort$T1_BDITot)
hist(Suicideshort$T2_BDITot)
hist(Suicideshort$T3_BDITot)
```
They are not...
This suggests we might want to use a different estimator that is more robust to non-normality. Let's try the MLM estimator.
```{r}
SEMRandSlopesMLM <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot'
SEMRandSlopesfitMLM <- growth(SEMRandSlopesMLM, orthogonal = TRUE, estimator = "MLM", data = Suicideshort)
summary(SEMRandSlopesfitMLM)
```

We don't see many differences between ML and MLM, likely because MLM is an extension of ML.
We could try another one: WLSMVS. This type of estimator is best for categorical data, so it may not be the best fit for my non-normal, continuous variable.
```{r}
SEMRandSlopesWLSMVS <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot'
SEMRandSlopesfitWLSMVS <- growth(SEMRandSlopesWLSMVS, orthogonal = TRUE, estimator = "WLSMVS", data = Suicideshort)
summary(SEMRandSlopesfitWLSMVS)
```
We see the slope and intercept vary slightly, but the variances are different.

#########BOOTSTRAPPING##############

If you want to bootstrap:
```{r}
SEMRandSlopesBoot <- ' i =~ 1*T1_BDITot + 1*T2_BDITot + 1*T3_BDITot 
                  s =~ 0*T1_BDITot + 1*T2_BDITot + 2*T3_BDITot
                  s~~0*s'
SEMRandSlopesfitBoot <- growth(SEMRandSlopesBoot, orthogonal = TRUE, estimator = "MLM", data = Suicideshort)
SEMRandSlopesBoot <- bootstrapLavaan(SEMRandSlopesfitBoot, R = 1000L, type = "nonparametric")
summary(SEMRandSlopesfitBoot)
```
Is it supposed to say somewhere in the output that this was bootstrapped?
